Human capital mediates natural selection in contemporary humans

### 2022-05-14 why-natural-selection.Rmd
Fix abstract error; finally order guides right and justify them; tweak colours

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores that predict lower earnings, education and health also predict lower fertility. Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with the economic theory of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "why-natural-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

  if (conf_int) {
    confint_scale <- scale_linetype_manual(
                       name   = "", 
                       values = c("95% c.i. uncorrected" = 1),
                       guide  = guide_legend(order = 1)
                     )
    confint_geom_segment <- confint_geom_segment()
  } else {
    confint_scale <- NULL
    confint_geom_segment <- NULL
  }

  n_regs <- as.double(n_regs)
  shape_values <- c(16, 15, 17, 18, 20)[1:n_cats]
  if (fill_direction == -1) shape_values <- rev(shape_values)
  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ...)) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, alpha = 0.9, aes(color = {{fill_col}})) +
    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/{{n_regs}})) +
    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
    scale_shape_manual(values = shape_values) +
    my_labs() +
    my_theme() +
    confint_scale +
    scale_alpha_discrete(range = c(1, 0), 
                           guide = guide_legend(
                             order = 0,
                             override.aes = list(
                               shape = c(1, 16),
                               color = "black",
                               alpha = 1,
                               size = 1.8
                             )
                         )) 
}

my_theme <- function (...) theme(axis.text.y = element_text(size = 7), ...)

@@

  scale <- if (n <= 2) {
    dots <- list(...)
    dots$direction <- NULL
    do.call(scale_colour_manual, c(list(values = fill_colors[1:n],
                                        aesthetics = aesthetics), 
                                   dots))
  } else {
    # scale <- scale_color_viridis_d(aesthetics = aesthetics, option = "D",
    #                                  direction = direction, end = 0.6, ...)
    if (direction != -1) fill_colors <- rev(fill_colors)
    scale_color_manual(aesthetics = aesthetics, 
                         values = shades::gradient(fill_colors,
                                                     steps = n), 
                         ...) 
  }

@@

res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/23` = sig 
      ) %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/23`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i.",
                        ),
                        width = 0.3,
                        color = "grey50"
                      ) +
        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "orange2")) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i." = 1))

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) plots mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
red bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups.
Overall, natural selection increases the correlation of polygenic scores with
income for `r n_increases` out of 33 polygenic scores, with a median percentage
increase of `r median_pct_change`% in the respondents' generation (Appendix
Table \@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores
are important for understanding social structure and mobility
[@belsky2018genetic; @Rimfeld_2018; @harden2021genetic], then these increases
are substantive. Similarly, since many polygenic scores are predictive of
disease risk, they could potentially increase health inequalities. In general,
the evolutionary history of anatomically modern humans is related to disease
risk [@benton2021influence]; understanding the role of contemporary natural
selection may help researchers to map the genetic architecture of health
disparities.

>>>>

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores that predict higher earnings, education and health also predict lower fertility. Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with the economic theory of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "why-natural-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

  if (conf_int) {
    confint_scale <- scale_linetype_manual(
                       name   = "", 
                       values = c("95% c.i. uncorrected" = 1),
                       guide  = guide_legend(order = 3)
                     )
    confint_geom_segment <- confint_geom_segment()
  } else {
    confint_scale <- NULL
    confint_geom_segment <- NULL
  }

  my_shape_fill_guide <- guide_legend(order = 1)
  n_regs <- as.double(n_regs)
  shape_values <- c(16, 15, 17, 18, 20)[1:n_cats]
  if (fill_direction == -1) shape_values <- rev(shape_values)
  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ...)) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, alpha = 0.9, aes(color = {{fill_col}})) +
    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/{{n_regs}})) +
    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction, 
                    guide = my_shape_fill_guide) +
    scale_shape_manual(values = shape_values, guide = my_shape_fill_guide) +
    my_labs() +
    my_theme(legend.justification = "top") +
    confint_scale +
    scale_alpha_discrete(range = c(1, 0), 
                           guide = guide_legend(
                             order = 2,
                             override.aes = list(
                               shape = c(1, 16),
                               color = "black",
                               alpha = 1,
                               size = 1.8
                             )
                         )) 
}

my_theme <- function (...) {
  theme(
          axis.text.y = element_text(size = 7), 
          ...
        )
}

@@

  scale <- if (n <= 2) {
    dots <- list(...)
    dots$direction <- NULL
    do.call(scale_colour_manual, c(list(values = fill_colors[1:n],
                                        aesthetics = aesthetics), 
                                   dots))
  } else {
    # scale <- scale_color_viridis_d(aesthetics = aesthetics, option = "D",
    #                                  direction = direction, end = 0.6, ...)
    fill_colors <- c(fill_colors[1], "skyblue", fill_colors[2])
    if (direction != -1) fill_colors <- rev(fill_colors)
    scale_color_manual(aesthetics = aesthetics, 
                         values = shades::gradient(fill_colors,
                                                     steps = n, space = "Lab"), 
                         ...) 
  }

@@

res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/23` = sig 
      ) %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/23`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i. uncorrected",
                        ),
                        width = 0.2,
                        color = "grey50"
                      ) +
        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "blue3")) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i. uncorrected" = 1))

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) plots mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
red bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment is stronger at the bottom of the income distribution, and
this increases the differences between groups. Overall, natural selection
increases the correlation of polygenic scores with income for `r n_increases`
out of 33 polygenic scores, with a median percentage increase of 
`r median_pct_change`% in the respondents' generation (Appendix Table
\@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores are
important for understanding social structure and mobility [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive.
Similarly, since many polygenic scores are predictive of disease risk, they
could potentially increase health inequalities. In general, the evolutionary
history of anatomically modern humans is related to disease risk
[@benton2021influence]; understanding the role of contemporary natural selection
may help researchers to map the genetic architecture of health disparities.
